Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-23 demystify.tex
updates

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{chap:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors. Just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}. Don't use AI to write your SOP. Not that hard to raise suspicion. Though you can ask AI to check your grammar and spelling.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some research and mention why you want to spend the next 5--7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}). Do your homework and mention profs who are till active in research. 
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). CS academics like using LaTeX (common way to write our papers and other documents), so write your SOP using LaTeX (with Times or default font, 11pt, and 1-inch margin).
\end{itemize}

@@

Many universities now require a diversity, equity, and inclusion (DEI) statement as part of the application. Students, especially international ones, often struggle with what to write, especially when they are not from a minority group or do not have much experience in DEI.

@@

Having an external and major fellowship can significantly improve your chances of getting into a good PhD program (\autoref{sec:fellowships}).  The National Science Foundation Graduate Research Fellowship Program (NSF GRFP) and the Department of Defense National Defense Science and Engineering Graduate Fellowship (DoD NDSEG) are two of the most prestigious fellowships for CS PhD students who are US citizens or permanent residents.  
These fellowship provide multiple years of full financial support, which includes a stipend and tuition. 
It is rumored that several programs would revert their admission decision on a student if they get one of these fellowships.

The \href{https://www.nsfgrfp.org/}{GRFP} and \href{https://www.ndsegfellowships.org/}{NDSEG} website have detailed information on the application process and eligibility requirements. This section focuses on how your application is evaluated and tips on improving your application. Note that while these are specific GRFP and NDSEG, the general advice can be applied to other major fellowships (those for non-US citizens or permanent residents).

@@

\item \textbf{Too Technical.} Remember that you're writing to a general audience in CS, not experts in your field.  While NSF tries to match reviewers with your field, CS fields are simply too broad and you might get a reviewer in the field but not in your specific area or familiar with the problem you're working on. 
So avoid technical jargon and explain things in simple terms. If the reviewer cannot understand your proposal, they will likely give you a low score.

You should ask your professor or someone who's familiar with writing NSF proposals to review your reseach plan.  They can tell you if it is too technical and help you simplify it.

@@

Finally, just as with the NSF GRFP and even PhD SOPs, ask your LoR writers and adviser to look at your writing/essays. You are new to the field and research writing, and your ideas might not be accurate or have already been explored. It's better that they tell you now than for you to find out later from reviewers.

>>>>

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{chap:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors. Just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}. Don't use AI to write your SOP. Not that hard to raise suspicion. Though you can ask AI to check your grammar and spelling.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some research and mention why you want to spend the next 5--7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}). Do your homework and mention profs who are till active in research. 
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). Don't use left alignment (seems to be default in Words) as it is hard to read.
CS academics like using \textbf{LaTeX} (common way to write our papers and other documents), so write your SOP using LaTeX (with Times or default font, 11pt, and 1-inch margin).
\end{itemize}

@@

Some universities require a diversity, equity, and inclusion (DEI) statement as part of the application. Students, especially international ones, often struggle with what to write, especially when they are not from a minority group or do not have much experience in DEI.

@@

Having an external and major fellowship can significantly relieve the financial burden of your PhD and improve your chances of getting into a good PhD program (\autoref{sec:fellowships}). 
Among the most prestigious fellowships for CS PhD students who are US citizens or permanent residents are the National Science Foundation Graduate Research Fellowship Program (NSF GRFP) and the Department of Defense National Defense Science and Engineering Graduate Fellowship (DoD NDSEG). These fellowships provide multiple years of full financial support, which includes a stipend and tuition.

The \href{https://www.nsfgrfp.org/}{GRFP} and \href{https://www.ndsegfellowships.org/}{NDSEG} websites have detailed information on the application process and eligibility requirements. This section focuses on how your application is evaluated and tips on improving it. Note that while these are specific GRFP and NDSEG, the general advice can be applied to other major fellowships (those for non-US citizens or permanent residents).

@@

\item \textbf{Too Technical or Narrow.} Remember that you're writing to a general audience in CS, not experts in your field.  While NSF tries to match reviewers with your field, CS fields are simply too broad and you might get a reviewer in the field but not in your specific area or familiar with the problem you're working on.
So avoid technical jargon and explain things in simple terms. If the reviewer cannot understand your research plan, they will likely give you a low score.

You should ask your professor or someone who's familiar with writing NSF proposals to review what you wrote.  They can tell you if it is too technical or narrow and help you improve it.

@@

\section{Common Tips for Both Fellowships}

\begin{itemize}
  \item \textbf{Use LaTeX to write your research plan and essays}.  This is the standard in CS and engineering and will make your application look more professional.  Moreover, do not use the default Word settings, e.g., left justification, as it is not standard in CS and engineering and difficult to read.

  \item \textbf{Ask your LoR writers and adviser to look at your writing/essays}. You are new to the field and research writing, and your ideas might not be accurate or have already been explored. It's better that they tell you now than for you to find out later from reviewers.
\end{itemize}
